\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-13 k-planar.tex

  \todo[inline]{%
    DW:Does this work for multigraph near-triangulations? \newline
    PM: Yes, it only requires that the inner faces have three sides and that the outer face is bounded by a simple cycle.  The proof goes through without change.  Look at the subgraph $X$ of the dual that contains only the edges dual to red/blue edges.  In our setting, the outer face has degree $1$ in $X$, which is odd.  Any bichromatic inner face has degree $0$ or degree $2$ in $X$.  There has to be another vertex of odd degree somewhere, which can only come from a trichromatic inner face.
  }

@@

\begin{thm}
\label{CliqueFaceProductStructure}
Let $G$ be a 2-connected plane multigraph, in which every facial cycle has length in $\{3,\dots,d\}$ for some fixed integer $d\geq 3$. Then $\widehat{G}$ is contained in $H \boxtimes P \boxtimes K_{ d +3( \floor{d/2}-1) }$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

@@

\begin{proof}[Proof of \cref{CliqueFaceProductStructure}]
Let $G$ be a 2-connected plane multigraph, in which every facial cycle has length in $\{3,\dots,d\}$. Let $C$ be the outer-cycle of $G$. Let $G^+$ be obtained from $G$ by adding a new vertex $r$ adjacent to every vertex on the outerface of $G$. So $G^+$ is a 2-connected plane multigraph and $G=G^+\blah{C}$. Let $T$ be a BFS spanning tree of $G^+$ rooted at $r$. Let $\mathcal{L}=(L_0,L_1,\dots)$ be the corresponding BFS layering. By \cref{NewLemma}, $G$ has a partition $\PP$ into tripods with respect to $T$ such that $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ for which every part of $\PP$ that intersects $F$ is in $B$. Since $G$ is a spanning subgraph of $\widehat{G}$, we may consider $\PP$ to be a partition of $\widehat{G}$, and $G/\PP \subseteq \widehat{G}/\PP$. Consider an edge $XY$ of $\widehat{G}/\PP$. By the definition of $\widehat{G}$, there is a facial cycle $F$ of $G$ such that both $X$ and $Y$ intersect $F$. By assumption, there is a bag $B$ of $\mathcal{T}$ such that every part of $\PP$ that intersects $F$ is in $B$. In particular, $X,Y\in B$. Hence $\TT$ is also a tree-decomposition of $\widehat{G}/\PP$. For each edge $vw$ of $\widehat{G}$, if $v\in L_i$ and $w\in L_j$ then $|i-j|\leq\floor{\frac{d}{2}}$. Grouping blocks of $\floor{\frac{d}{2}}$ consecutive layers in $\mathcal{L}$ gives a layering $\mathcal{L}'$ of $\widehat{G}$. Observe that each tripod in $\PP$ has at most $d+3( \floor{\frac{d}{2}}-1)$ vertices in each set of $\floor{\frac{d}{2}}$ layers of $\mathcal{L}$. Hence $\mathcal{L}'$ has layered width $d+3( \floor{\frac{d}{2}}-1)$ with respect to $\PP$. By \cref{PartitionProduct}, $\widehat{G}$ is contained in $H \boxtimes P \boxtimes K_{ d + 3( \floor{d/2} -1 )}$ where $H:=\widehat{G}/\PP$, which has treewidth 3.
\end{proof}

@@

\begin{cor}
	\label{New1Planar}
	Every 1-planar graph is contained in $H \boxtimes P \boxtimes K_8$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{cor}

>>>>

%  \todo[inline]{%
%    DW:Does this work for multigraph near-triangulations? \newline
%    PM: Yes, it only requires that the inner faces have three sides and that the outer face is bounded by a simple cycle.  The proof goes through without change.  Look at the subgraph $X$ of the dual that contains only the edges dual to red/blue edges.  In our setting, the outer face has degree $1$ in $X$, which is odd.  Any bichromatic inner face has degree $0$ or degree $2$ in $X$.  There has to be another vertex of odd degree somewhere, which can only come from a trichromatic inner face.\newline
%    DW: great. This is basically the same as the proof of the Hex Lemma.
%  }

@@

\begin{thm}
\label{CliqueFaceProductStructure}
Let $G$ be a 2-connected plane multigraph, in which every facial cycle has length in $\{3,\dots,d\}$ for some fixed integer $d\geq 3$. Then $\widehat{G}$ is contained in $H \boxtimes P \boxtimes K_{ d +3 \floor{d/2}-3 }$ for some planar graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

@@

\begin{proof}[Proof of \cref{CliqueFaceProductStructure}]
Let $G$ be a 2-connected plane multigraph, in which every facial cycle has length in $\{3,\dots,d\}$. Let $C$ be the outer-cycle of $G$. Let $G^+$ be obtained from $G$ by adding a new vertex $r$ adjacent to every vertex on the outerface of $G$. So $G^+$ is a 2-connected plane multigraph and $G=G^+\blah{C}$. Let $T$ be a BFS spanning tree of $G^+$ rooted at $r$. Let $\mathcal{L}=(L_0,L_1,\dots)$ be the corresponding BFS layering. By \cref{NewLemma}, $G$ has a partition $\PP$ into tripods with respect to $T$ such that $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ for which every part of $\PP$ that intersects $F$ is in $B$. Since $G$ is a spanning subgraph of $\widehat{G}$, we may consider $\PP$ to be a partition of $\widehat{G}$, and $G/\PP \subseteq \widehat{G}/\PP$. Consider an edge $XY$ of $\widehat{G}/\PP$. By the definition of $\widehat{G}$, there is a facial cycle $F$ of $G$ such that both $X$ and $Y$ intersect $F$. By assumption, there is a bag $B$ of $\mathcal{T}$ such that every part of $\PP$ that intersects $F$ is in $B$. In particular, $X,Y\in B$. Hence $\TT$ is also a tree-decomposition of $\widehat{G}/\PP$. For each edge $vw$ of $\widehat{G}$, if $v\in L_i$ and $w\in L_j$ then $|i-j|\leq\floor{\frac{d}{2}}$. Grouping blocks of $\floor{\frac{d}{2}}$ consecutive layers in $\mathcal{L}$ gives a layering $\mathcal{L}'$ of $\widehat{G}$. Observe that each tripod in $\PP$ has at most $d+3\floor{\frac{d}{2}}-3$ vertices in each set of $\floor{\frac{d}{2}}$ layers of $\mathcal{L}$. Hence $\mathcal{L}'$ has layered width $d+3( \floor{\frac{d}{2}}-1)$ with respect to $\PP$. By \cref{PartitionProduct}, $\widehat{G}$ is contained in $H \boxtimes P \boxtimes K_{d+3\floor{d/2}-3}$ where $H:=\widehat{G}/\PP$, which has treewidth 3. \note{DW}{This proof needs to be changed to show that $G/\PP = \widehat{G}/\PP$, so $H$ is a minor of $G$ and is thus planar.}
\end{proof}

@@

\begin{cor}
	\label{New1Planar}
	Every 1-planar graph is contained in $H \boxtimes P \boxtimes K_7$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{cor}
